Supplemental Information: Aligning marine species range data to better serve science and conservation

### 2016-09-01 clip_depth/coral_maps.Rmd
some updates to methods and res/disc re: corals and aquamaps clipping
incorporating Halley's suggestions; updated figs to match PLOS-ONE requirements; bringing in Jamie's supplemental figures

@@

save_tmap(map_coral_comp,
          filename = file.path(dir_fig, 'fig3a_coral_map.png'), 
          width = 9, height = 4.2, units = 'cm', dpi = 600)

@@

violplot1 <- ggplot(spp_coral_clip, aes(x = 'sm_range_raw', y = shallow_pct)) +
  ggtheme_plot +
  # geom_violin(fill = 'grey80', alpha = .5,
  #             draw_quantiles = c(.25, .5, .75)) +
  # geom_point(position = position_jitter(w = .3), 
  #            size = 0.25,
  #            color = 'black', alpha = 1) +
  geom_violin(fill = NA, size = 0.1, weight = 0.5,
              draw_quantiles = c(seq(.005, .05, .005), seq(.05, .95, .02), seq(.95, 1.0, .005)),
              color = 'black') +
  geom_violin(fill = NA, size = 1,
              draw_quantiles = NULL,
              color = 'white') +
  geom_hline(color = 'grey30', yintercept = 100) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_discrete(labels = element_blank()) +
  theme(axis.title.x = element_blank()) +
  labs(y = 'IUCN coral range: (<= 200 m) / (total)')
print(violplot1)

@@

### Plot the new quad graph
clipped_quads <- ggplot(iucn_coral_fixed %>%
                      filter(!iucn_sid %in% (iucn_coral_fixed %>% 
                                               filter(perc > 100) %>% 
                                               .$iucn_sid)),
                    aes(x = a_ratio, y = d_perc, group = iucn_sid)) +
  ggtheme_plot + 
  ### color the quadrant backgrounds:
  annotate("rect", xmin = 54.5, xmax = 100, ymin = 63, ymax = 100,
           alpha = .3, fill= "#4dac26")  +
  annotate("rect", xmax = 54.5, xmin =   0, ymin = 63, ymax = 100,
           alpha = .3, fill= "#b8e186") +
  annotate("rect", xmin = 54.5, xmax = 100, ymax = 63, ymin =   0,
           alpha = .3, fill= "#f1b6da") +
  annotate("rect", xmax = 54.5, xmin =   0, ymax = 63, ymin =   0,
           alpha = .3, fill= "#d01c8b") +
  ### add grid lines
  geom_vline(xintercept = break_nums, size = 0.25, color = 'white', alpha = .5) +
  geom_hline(yintercept = break_nums, size = 0.25, color = 'white', alpha = .5) +

@@

ggsave(plot = clipped_quads,
       filename = file.path(dir_fig, 'fig3b_scatter_quadmap_corals.png'), device = 'png',
       width = 17.8, height = 9, units = 'cm', dpi = 600)

@@

ggsave(plot = barchart_coral_quads,
       filename = file.path(dir_fig, 'fig3c_barchart_coral_quads.png'),
       width = 17.8, height = 4, units = 'cm', dpi = 600)

>>>>

coral_cells_iucn <- read_csv(file.path(dir_git, 'clip_depth', 'coral_cells_iucn.csv'))

@@

save_tmap(map_coral_comp,
          filename = file.path(dir_fig, 'coral_map.png'), 
          width = 19, height = 8, units = 'cm', dpi = 600)

@@

@@

### Plot the new quad graph
clipped_quads <- ggplot(iucn_coral_fixed %>%
                      filter(!iucn_sid %in% (iucn_coral_fixed %>% 
                                               filter(d_perc > 100) %>% 
                                               .$iucn_sid)),
                    aes(x = a_ratio, y = d_perc, group = iucn_sid)) +
  ggtheme_plot + 
  ### color the quadrant backgrounds:
  annotate("rect", xmin = 54.5, xmax = 100, ymin = 63, ymax = 100,
           alpha = .3, fill= "#4dac26")  +
  annotate("rect", xmax = 54.5, xmin =   0, ymin = 63, ymax = 100,
           alpha = .3, fill= "#b8e186") +
  annotate("rect", xmin = 54.5, xmax = 100, ymax = 63, ymin =   0,
           alpha = .3, fill= "#f1b6da") +
  annotate("rect", xmax = 54.5, xmin =   0, ymax = 63, ymin =   0,
           alpha = .3, fill= "#d01c8b") +
  ### add grid lines
  geom_vline(xintercept = break_nums, size = 0.25, color = 'white', alpha = .5) +
  geom_hline(yintercept = break_nums, size = 0.25, color = 'white', alpha = .5) +

@@

ggsave(plot = clipped_quads,
       filename = file.path(dir_fig, 'si_scatter_quadmap_corals.png'), device = 'png',
       width = 19, height = 9, units = 'cm', dpi = 600)

@@

ggsave(plot = barchart_coral_quads,
       filename = file.path(dir_fig, 'barchart_coral_quads.png'),
       width = 19, height = 4, units = 'cm', dpi = 600)

### 2016-09-01 data_explore/data_explore.Rmd
incorporating Halley's suggestions; updated figs to match PLOS-ONE requirements; bringing in Jamie's supplemental figures

print(bar_taxa)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_all.png'), 
       width = 17.8, height = 7, units = 'cm', dpi = 600)

@@

ggsave(file.path(dir_fig, 'scatter_quadmap.png'),
       width = 17.8, height = 9, units = 'cm', dpi = 600)

@@

ggsave(plot = barchart_spp_gp_quads,
       filename = file.path(dir_fig, 'barchart_spp_gp_quads.png'),
       width = 17.8, height = 7, units = 'cm', dpi = 600)

>>>>

print(bar_taxa)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_all.png'), 
       width = 19.05, height = 7, units = 'cm', dpi = 600)

@@

ggsave(file.path(dir_fig, 'scatter_quadmap.png'),
       width = 19.05, height = 9, units = 'cm', dpi = 600)

@@

ggsave(plot = barchart_spp_gp_quads,
       filename = file.path(dir_fig, 'barchart_spp_gp_quads.png'),
       width = 19.05, height = 7, units = 'cm', dpi = 600)

### 2016-09-01 data_explore/data_explore_quadmaps.Rmd
incorporating Halley's suggestions; updated figs to match PLOS-ONE requirements; bringing in Jamie's supplemental figures

library(ggplot2)
library(maps)
library(readr)
library(data.table)
library(raster)
library(tidyr)
library(dplyr)
library(stringr)

dir_N <- c('Windows' = '//neptune.nceas.ucsb.edu/data_edit',
           'Darwin'  = '/Volumes/data_edit',
           'Linux'   = '/var/data/ohi')[[ Sys.info()[['sysname']] ]]

@@

dir_anx <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO')
dir_fig <- file.path(dir_git, 'figures')
dir_data <- file.path(dir_git, 'data')

@@

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN subpopulations
iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_2015.csv'), 
                           col_types = 'cddddc', progress = TRUE) %>%
  as.data.frame() %>%
  filter(presence != 5) ### removes extinct subpopulations

am_spp_cells   <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/am_spp_cells_2015.csv'), progress = TRUE)

### LOICZID cell ID numbers - used in plotting rasters
loiczid_raster_file  <- file.path(dir_git, 'shiny/data/loiczid_raster.grd')
loiczid_raster       <- raster(loiczid_raster_file)
names(loiczid_raster) <- 'loiczid'
```

@@

### mongo plot time
scatter_quadplot <- ggplot(spp_list,
                           aes(x = area_ratio, 
                               y = sm_perc, 
                               color = popn_category)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_align_mean, xmax = 100, 
                     ymin = dist_align_mean, ymax = 100, 
             alpha = .3, 
             fill= "#4dac26")  + 
    annotate("rect", xmax = area_align_mean, xmin =   0, 
                     ymin = dist_align_mean, ymax = 100, 
             alpha = .3, 
             fill= "#b8e186") + 
    annotate("rect", xmin = area_align_mean, xmax = 100, 
                     ymax = dist_align_mean, ymin =   0, 
             alpha = .3, 
             fill= "#f1b6da") + 
    annotate("rect", xmax = area_align_mean, xmin =   0, 
                     ymax = dist_align_mean, ymin =   0, 
             alpha = .3, 
             fill= "#d01c8b") + 
  geom_point(alpha = .5)

@@

``` {r plot_spp_from_quadrants, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
### define representative species from each quadrant, for over-plotting
spp_quads <- c('q1' = 'Kajikia albida', 
               'q2' = 'Conus episcopatus',
               'q3' = 'Conus magnificus', 
               'q4' = 'Acanthurus nigroris')
am_thresh <- 0

for (i in 1:length(spp_quads)) { # i = 1
  spp <- spp_list %>% filter(sciname == spp_quads[i])
  spp_quad <- names(spp_quads)[i]

  spp_plot <- plot_rangemap(spp)
  spp_plot <- spp_plot + 
    labs(title = element_blank()) +
    theme(legend.position = 'none')

  print(spp_plot)
  plotfile <- file.path(dir_fig, 'spp_maps', 
                        sprintf('%s_%s.png', 
                                spp_quad, 
                                tolower(str_replace(spp_quads[i], ' ', '_'))))
  ggsave(plotfile, height = 4.5, width = 8.7, units = 'cm')
  }

>>>>

library(ggplot2)
library(tmap)
library(maptools)
data(wrld_simpl)
library(maps)
library(readr)
library(data.table)
library(raster)
library(tidyr)
library(dplyr)
library(stringr)

dir_M <- c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
           'Darwin'  = '/Volumes/ohi',    ### connect (cmd-K) to smb://mazu/ohi
           'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

@@

dir_anx <- file.path(dir_M, 'git-annex/globalprep/spp_ico')
dir_fig <- file.path(dir_git, 'figures')
dir_data <- file.path(dir_git, 'data')

@@

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN subpopulations
iucn_spp_cells <- read_csv(file.path(dir_anx, 'vAM_IUCN/iucn_spp_cells_2015.csv'), 
                           col_types = 'cddddc', progress = TRUE) %>%
  filter(presence != 5) ### removes extinct subpopulations

am_spp_cells   <- read_csv(file.path(dir_anx, 'vAM_IUCN/am_spp_cells_2015.csv'), progress = TRUE)

### LOICZID cell ID numbers - used in plotting rasters
loiczid_raster_file  <- file.path(dir_git, 'data/rasters/loiczid_raster.tif')
loiczid_raster       <- raster(loiczid_raster_file)
names(loiczid_raster) <- 'loiczid'
```

@@

### mongo plot time
scatter_quadplot <- ggplot(spp_list,
                           aes(x = area_ratio, 
                               y = sm_perc, 
                               color = popn_category)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_align_mean, xmax = 100, 
                     ymin = dist_align_mean, ymax = 100, 
             alpha = .3, 
             fill= "#4dac26")  + 
    annotate("rect", xmax = area_align_mean, xmin =   0, 
                     ymin = dist_align_mean, ymax = 100, 
             alpha = .3, 
             fill= "#b8e186") + 
    annotate("rect", xmin = area_align_mean, xmax = 100, 
                     ymax = dist_align_mean, ymin =   0, 
             alpha = .3, 
             fill= "#f1b6da") + 
    annotate("rect", xmax = area_align_mean, xmin =   0, 
                     ymax = dist_align_mean, ymin =   0, 
             alpha = .3, 
             fill= "#d01c8b") + 
  geom_point(alpha = .5, size = 1)

@@

``` {r tmap_species_maps_fxns}

get_spp_map_df <- function(species, am_cutoff = 0){ ### species <- spp_list$sciname[1]
  message('in get_spp_map_df()')
  ### am_cutoff is currently ignored

  spp_id <- spp_list %>%
    filter(sciname == species) %>%
    dplyr::select(am_sid, iucn_sid, sciname) %>%
    distinct()

  iucn_spp_map <- iucn_spp_cells %>%
    filter(iucn_sid %in% spp_id$iucn_sid) %>%
    group_by(loiczid) %>%        
    summarize(iucn_sid = first(iucn_sid))
      ### The group_by() and summarize() are to eliminate duped cells (e.g. 
      ### one am_sid matching two iucn_sids)

  am_spp_map   <- am_spp_cells %>%
    filter(am_sid == spp_id$am_sid)

  spp_map_df <- full_join(iucn_spp_map, am_spp_map, by = 'loiczid') %>%
    mutate(am_pres = ifelse(is.na(am_sid), NA, 1),
           iucn_pres = ifelse(is.na(iucn_sid), NA, 1)) %>%
    dplyr::select(loiczid, am_pres, iucn_pres) %>%
    mutate(both_pres = ifelse(!is.na(iucn_pres), ### IUCN presence, so...
                              ifelse(is.na(am_pres), 2, 3), ### if no AM, assign 2; if AM, assign 3. 
                              am_pres))          ### no IUCN, so assign 1 for AM and NA for none

  return(spp_map_df)
}

get_rast <- function(spp_map_df, type) {
  message('in get_rast()')
  rast_obj  <-  subs(loiczid_raster, spp_map_df, 
                     by    = 'loiczid', 
                     which = paste0(type, '_pres'), 
                     subsWithNA = TRUE)

  rast_obj_shift <- rast_obj %>%
    raster::shift(360)

  rast_obj <- raster::merge(rast_obj, rast_obj_shift)
  return(rast_obj)
}

assemble_map <- function(map_rast, spp) {
  message('in assemble_map()')
  map_obj <- tm_shape(map_rast) +
    tm_raster(palette = 'Spectral',
              colorNA = NULL,
              title = spp,
              alpha = 1) +
    tm_shape(World) +
      tm_polygons() + 
    tm_layout(basemaps = "Esri.WorldTopoMap", title.position = 'TOP', legend.outside = TRUE, attr.outside = TRUE)

  return(map_obj)
}

wrld_simpl2 <- wrld_simpl %>%
  rgeos::gUnion(wrld_simpl %>% shift(360))

fao_polys <- rgdal::readOGR(dsn = file.path(path.expand(dir_git), 'clip_fao/fao_rgns'),
                            layer = 'fao_rgns2')

``` {r plot_spp_from_quadrants, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}

spp_quads <- c('q1' = 'Kajikia albida',       
               'q2' = 'Conus episcopatus',
               'q3' = 'Conus magnificus', 
               'q4' = 'Acanthurus nigroris')

spp_extents <- list(c(-150,  60, -75, 75), # 210 deg x 150 deg
                    c(  30, 240, -75, 75), # 210 deg
                    c(  90, 300, -75, 75), # 210 deg
                    c(  30, 240, -75, 75)) # 210 deg

spp_map_list <- vector('list', length = 4)

for (i in 1:length(spp_quads)) { # i = 1
  species = spp_quads[i]

  spp <- spp_list %>% filter(sciname == spp_quads[i])
  spp_quad <- names(spp_quads)[i]

  spp_map_df <- get_spp_map_df(species = spp_quads[i])

  spp_rast <- get_rast(spp_map_df, type = 'both') %>%
    crop(extent(spp_extents[[i]]))

  fao_polys_crop <- crop(fao_polys,   extent(spp_extents[[i]]))
  world_crop     <- crop(wrld_simpl2, extent(spp_extents[[i]]))

  # colvec <- c('AquaMaps' = '#76A94C', 'Both' = '#e18258', 'IUCN' = '#9168a9')

  if (i == 1) {
    spp_map <- tm_shape(spp_rast) +
      tm_raster(palette = c('#76A94C', '#9168a9', '#e18258'),
                title  = 'Dataset',
                labels = c('AquaMaps', 'IUCN', 'both'),
                colorNA = '#f0f8ff',
                showNA = FALSE,
                alpha = 1)
  } else {
    spp_map <- tm_shape(spp_rast) +
      tm_raster(palette = c('#76A94C', '#9168a9', '#e18258'),
                legend.show = FALSE,
                colorNA = '#f0f8ff',
                alpha = 1)
  }

  spp_map <- spp_map +
    tm_shape(fao_polys_crop) +
      tm_borders(col = 'grey70', lwd = 0.25, alpha = .3) +
    tm_shape(world_crop) +
      tm_polygons(border.col = 'grey45', col = 'grey40', lwd = 0.25) + 
    tm_layout(title = species, title.size = .5,
              frame = FALSE,
              legend.text.size = .35,
              legend.title.size = .45,
              # title.position = 'TOP', 
              legend.outside = FALSE,
              legend.position = c('right', 'bottom'),
              legend.bg.color = '#f0f8ff',
              legend.bg.alpha = .5,
              attr.outside = TRUE)

  spp_map_list[[i]] <- spp_map

  save_tmap(spp_map, file.path(dir_fig, sprintf('spp_maps/spp_map_%s_%s.png', 
                                                spp_quad, 
                                                str_replace(tolower(species), ' ', '_'))), 
            width = 9.4, height = 6.5, units = 'cm', dpi = 600)

}

# multiplot <- grid.arrange(spp_map_list[[2]],
#                           spp_map_list[[1]],
#                           spp_map_list[[4]],
#                           spp_map_list[[3]],
#                           ncol = 2, main = "Main title")

### 2016-09-01 manuscript-main.Rmd
incorporating Halley's suggestions; updated figs to match PLOS-ONE requirements; bringing in Jamie's supplemental figures

<!-- Full title	250 characters	Specific, descriptive, concise, and comprehensible to readers outside the field	 -->
<!--   Impact of Cigarette Smoke Exposure on Innate Immunity: A Caenorhabditis elegans Model -->
<!--   Solar Drinking Water Disinfection (SODIS) to Reduce Childhood Diarrhoea in Rural Bolivia: A Cluster-Randomized, Controlled Trial -->
<!-- Short title	70 characters	State the topic of the study	 -->
<!--   Cigarette Smoke Exposure and Innate Immunity -->
<!--   SODIS and Childhood Diarrhoea -->

_Author affiliation:_ <!--Include department, institution, and complete address, with the ZIP/postal code, for each author. Use superscripts to match authors with institutions. Authors are encouraged to supply their ORCID ID at submission-->

>>>>

Short title: "Aligning marine species range data"
<!-- Full title	250 characters	Specific, descriptive, concise, and comprehensible to readers outside the field	 -->
<!--   Impact of Cigarette Smoke Exposure on Innate Immunity: A Caenorhabditis elegans Model -->
<!--   Solar Drinking Water Disinfection (SODIS) to Reduce Childhood Diarrhoea in Rural Bolivia: A Cluster-Randomized, Controlled Trial -->
<!-- Short title	70 characters	State the topic of the study	 -->
<!--   Cigarette Smoke Exposure and Innate Immunity -->
<!--   SODIS and Childhood Diarrhoea -->

### 2016-09-01 ms_1_intro.Rmd
incorporating Halley's suggestions; updated figs to match PLOS-ONE requirements; bringing in Jamie's supplemental figures

Knowing where species exist and thrive is fundamental to the sciences of ecology, biogeography, and conservation, among many others.  This knowledge provides foundational information for 
 understanding species ranges and diversity,
 predicting species responses to human impacts and climate change,
 and managing and protecting species effectively.
<!--___A rich literature tackles the many dimensions of these questions.___-->

One major outcome of this body of science is the various compiled databases of species distribution maps. The two most comprehensive, widely-used global-scale repositories that predict marine species ranges throughout the world's oceans are AquaMaps [@kaschner_aquamaps:_2015] and range data from the International Union for Conservation of Nature (IUCN) [@iucn_iucn_2015].  Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each offers an understanding of the truth, with IUCN ranges relying primarily on expert opinion and AquaMaps on model predictions.  Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset: geographic range data such as IUCN range maps frequently introduce commission errors (inaccurate indication of presence), while species distribution models such as AquaMaps will likely introduce fewer commission errors at the expense of more omission errors (inaccurate indications of absence) [@rondinini_tradeoffs_2006].  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006; @jetz_ecological_2008].

These two spatial datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status [@halpern_index_2012; @halpern_patterns_2015; @selig_assessing_2013-1], evaluating global biodiversity patterns [@coll_biodiversity_2010; @martin_manual_2014; @pimm_biodiversity_2014; @kaschner_current_2011], predicting species range shifts [@garciamolinos_climate_2015], and setting conservation priorities [@klein_shortfalls_2015].  In most of these cases, the implications of choosing one versus the other of these datasets is not evaluated or discussed, yet strong conservation and management conclusions are drawn from the results of the studies. <!-- For example, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.-->

@@

Although relatively small in number, these overlap species present a unique opportunity to evaluate the two datasets overall.  As understanding of a given species improves, we expect that range maps, regardless of method, would become more accurate in predicting species presence and absence.  Close alignment between two independent range maps can therefore indicate higher confidence in species presence or absence than a single map.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align in both distribution and overall area; from the results we determine methodological issues that can introduce commission errors, and suggest methods to improve confidence in species range predictions.  In particular, we show how solving a simple problem in coral ranges in the IUCN range data dramatically ___improves agreement with Aquamaps data.___ - _does this make it sound like agreement is the goal? make it clear that agreement implies better confidence_

@@

We then reexamine a global analysis of gaps in protection afforded by marine protected areas (MPAs) [@klein_shortfalls_2015] - as a case study to explore the implications of prioritizing one data set over the other.  The results highlight possible consequences of different data use decisions on our understanding of marine biodiversity status and protection.

While we have identified some potential issues in each dataset, we cannot simply recommend one data set over the other.  Instead, we recommend simple methods to reduce incidence of errors in each dataset, improving confidence in species range predictions to better inform conservation management and policy decisions.

>>>>

Knowing where species exist and thrive is fundamental to the sciences of ecology, biogeography, and conservation.  This body of science has resulted in various compiled databases of species distribution maps which provide foundational information for understanding species ranges and diversity, predicting species responses to human impacts and climate change, and managing and protecting species effectively.  No species range dataset can claim to represent the "truth" of any species' spatial distribution, but rather each offers its own distinct and often divergent understanding of the "true" distribution.  Despite the uncertainty and variation among species range datasets, conservation managers and policy makers must act on conclusions drawn from these data.  It is critical to understand the differences in spatial range datasets and the implications of these differences on conservation decisions.

<!--implications of choosing one versus the other of these datasets is not evaluated or discussed, yet strong conservation and management conclusions are drawn from the results of the studies.-->

The two most comprehensive, widely-used global-scale repositories that predict marine species ranges throughout the world's oceans are AquaMaps, which rely primarily on model predictions [@kaschner_aquamaps:_2015], and range data from the International Union for Conservation of Nature (IUCN), which rely primarily on expert opinion [@iucn_iucn_2015].  Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset: geographic range data such as IUCN range maps frequently introduce commission errors (inaccurate indication of presence), while species distribution models such as AquaMaps will likely introduce fewer commission errors at the expense of more omission errors (inaccurate indications of absence) [@rondinini_tradeoffs_2006].  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006; @jetz_ecological_2008].

These two spatial datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status [@halpern_index_2012; @halpern_patterns_2015; @selig_assessing_2013-1], evaluating global biodiversity patterns [@coll_biodiversity_2010; @martin_manual_2014; @pimm_biodiversity_2014; @kaschner_current_2011], predicting species range shifts [@garciamolinos_climate_2015], and setting conservation priorities [@klein_shortfalls_2015].  In most of these cases, the implications of choosing one versus the other of these datasets is not evaluated or discussed, yet strong conservation and management conclusions are drawn from the results of the studies.

@@

Although relatively small in number, these overlap species present a unique opportunity to evaluate the two datasets overall.  As understanding of a given species improves, we expect that range maps, regardless of method, would become more accurate in predicting species presence and absence.  Close alignment between two independent range maps can therefore indicate higher confidence in species presence or absence than a single map.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align in both distribution and overall area.

Using this alignment framework, we explore methodological issues that can introduce errors, and suggest methods to improve confidence in species range predictions.  In particular, we show how explicitly using depth as a constraint on coral ranges in IUCN range data dramatically improves concurrence with AquaMaps data, ___and highlight a side effect of AquaMaps methodology that results in surprising species range boundaries.___

@@

As a case study, we then reexamine a global analysis of gaps in protection afforded by marine protected areas (MPAs) [@klein_shortfalls_2015] - as a case study to explore the implications of prioritizing one data set over the other.  The results highlight possible consequences of different data use decisions on our understanding of marine biodiversity status and protection.

While we have identified some potential issues in each dataset, we cannot simply recommend one data set over the other.  Instead, we recommend simple methods to account for incidence of errors in each dataset, improving confidence in species range predictions to better inform conservation management and policy decisions.

### 2016-09-01 ms_2_methods.Rmd
incorporating Halley's suggestions; updated figs to match PLOS-ONE requirements; bringing in Jamie's supplemental figures

The IUCN publishes species range maps developed by species experts.  These experts outline spatial boundaries of a given species' "limits of distribution" [@iucn_iucn_2015], based on observation records and informed by expert understanding of species' range and habitat preferences.  ___Fig SXXX___

In contrast, AquaMaps models species distribution based on environmental preferences, such as temperature, depth, and salinity, deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a global 0.5° grid to determine suitable habitat, resulting in a "probability of occurrence" for each species.  Of these, 1296 (5.7%) have been further refined through an expert review process [@kaschner_mapping_2006; @ready_predicting_2010]. ___Fig SXXX___

@@

__Comparison of taxonomic and regional distribution__:
To examine the overall taxonomic distribution across the spatial datasets, we grouped species by taxonomic class and data source, and determined the proportion of each class represented in each dataset.  To compare the spatial representation of the two datasets directly, we rasterized the IUCN species polygons to the same 0.5° grid as the AquaMaps species maps; we determined species presence within a grid cell as any non-zero overlap of a species polygon with the cell  ___(Fig SXXX)___.  For the AquaMaps dataset, we determined per-cell species count by including all species with non-zero probability of occurrence, to best approximate the "extent of occurrence" generally indicated by IUCN maps ___(Fig SXXX)___.

__Comparison of paired maps__:
Using genus and species binomials as a matching key, we identified "paired map" species - the subset of marine species that have range maps in both IUCN and AquaMaps current native distribution.  We used the taxize package [@REF] in R [@REF] to standardize species names and synonyms; for species with separate subpopulation maps in IUCN, we combined all subpopulations to create a single global population.   For each of these paired map species, we determined species presence within each spatial cell for each dataset using the same criteria outlined above.

@@

For each paired map species, $A_{small}$ and $A_{large}$ indicate the smaller and larger range representation (regardless of which dataset).  $A_{small \cap large}$ represents the amount of overlapping area between the two datasets.

Distribution alignment uses overlapping predictions of presence as means of identifying areas of higher confidence.  Area ratio provides a proxy for frequency of commission and/or omission errors.

To examine errors of commission related to depth, we selected corals as a case study due to their importance in supporting biodiversity as well as their dependence on photosynthesis.  Extracting data from the IUCN API [@REF?] we identified the depth limitations of each of the coral species mapped in the IUCN dataset.  We created a 200 m bathymetry raster from Natural Earth's 200 m bathymetry polygon [@natural_earth_XXX] ___(Fig SXXX)___ and masked our IUCN coral rasters to identify mapped coral presence below 200 m.  The resulting maps were again compared to the AquaMaps rasters to examine distribution alignment and area ratio.

AquaMaps uses Food and Agriculture Organization of the United Nations (FAO) Major Fishing Area [@REF] boundaries as a rough method of constraining species ranges to appropriate georegions.  To explore the impact of these artificial constraints on species range predictions, we chose to focus on boundaries defined by longitude, as they seem less likely to relate to ecological conditions than boundaries defined by latitude.  As an example, we identified all AquaMaps species whose eastern or western range limit between between 5° S and 25° N latitudes coincided with the large vertical boundary between FAO regions 71 and 77 at 175° W longitude.

To identify data-poor species, we used the ___robis???___ and ___rgbif???___ packages [@REFS] in R to identify known occurrences of each species; occurrences were averaged between the two occurrence databases as a proxy for data richness.

>>>>

The IUCN publishes species range maps developed by species experts.  These experts outline spatial boundaries of a given species' "limits of distribution" [@iucn_iucn_2015], based on observation records and informed by expert understanding of species' range and habitat preferences (Fig S1A)

In contrast, AquaMaps models species distribution based on envelopes of environmental preference, such as temperature, depth, and salinity; these preference envelopes are deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a global 0.5° grid to determine suitable habitat, resulting in a "probability of occurrence" for each species (Fig S1B).  To roughly constrain species ranges to appropriate georegions, the AquaMaps model uses Food and Agriculture Organization of the United Nations (FAO) Major Fishing Area [@REF] boundaries.   Of the resulting maps, 1296 (5.7%) have been further refined through an expert review process [@kaschner_mapping_2006; @ready_predicting_2010].

@@

__Comparison of taxonomic and regional distribution__:
To examine the overall taxonomic distribution across the spatial datasets, we grouped species by taxonomic class and data source, and determined the proportion of each class represented in each dataset.  To compare the spatial representation of the two datasets directly, we rasterized the IUCN species polygons to the same 0.5° grid as the AquaMaps species maps; we determined species presence within a grid cell as any non-zero overlap of a species polygon with the cell  ___(Fig S2)___.  For the AquaMaps dataset, we determined per-cell species count by including all species with non-zero probability of occurrence, to best approximate the "extent of occurrence" generally indicated by IUCN maps.

__Alignment framework for paired maps__: OR __Comparison of paired maps__:
Using genus and species binomials as a matching key, we identified "paired map" species, i.e. the subset of marine species that have range maps in both IUCN and AquaMaps current native distribution.  We used the taxize package [@REF] in R [@REF] to standardize species names and synonyms; for species with separate subpopulation maps in IUCN, we combined all subpopulations to create a single global population.   For each of these paired map species, we determined species presence within each spatial cell for each dataset using the same criteria outlined above.

@@

For each paired map species, $A_{small}$ and $A_{large}$ indicate the smaller and larger range representation (regardless of which dataset).  $A_{small \cap large}$ represents the amount of overlapping area between the two datasets.  Distribution alignment uses overlapping predictions of presence as means of identifying areas of higher confidence.  Area ratio provides a proxy for frequency of commission and/or omission errors.

__(some heading here? to separate from alignment)__
Corals provide a useful case study to examine differences in species range alignment.  They are an intensely studied taxonomic group whose importance in supporting biodiversity is undisputed; their dependence on photosynthesis provides habitat-limiting life history traits, and they don't tend to move.  Extracting data from the IUCN API [@REF?] we identified depth limitations of each coral species mapped in the IUCN dataset to verify that none are found below the photosynthetic depth limit of 200 m.  Using the same methodology as shown in Fig 2, we created a 200 m bathymetry raster from Natural Earth's 200 m bathymetry polygon [@natural_earth_XXX] and masked our IUCN coral rasters to identify mapped coral presence below 200 m.  The resulting maps were again compared to the AquaMaps rasters to examine distribution alignment and area ratio.

AquaMaps' reliance upon FAO zones to constrain species ranges seems likely to result in odd species range boundaries where ecologically suitable habitat intersects with human-defined borders.  To explore the impact of these artificial constraints on species range predictions, we chose to focus on boundaries defined by longitude, as they seem less likely to coincide with ecological differences than boundaries defined by latitude.  As an example, we identified all AquaMaps species whose eastern or western range limit between between 5° S and 25° N latitudes coincided with the large vertical boundary between FAO regions 71 and 77 at 175° W longitude.

The oceans abound with rare and poorly-understood species who offer sparse information on behavior and habitat preferences to inform models and expert opinion.  To identify such data-poor species within the set of paired map species, we used the ___robis???___ and ___rgbif???___ packages [@REFS] in R to identify known occurrences of each species; occurrences were averaged between the two occurrence databases as a proxy for data richness.
